---
output:
  github_document:
    pandoc_args: ["--wrap=none", "--from=markdown-smart"]
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  dev = "svg",
  warning = FALSE,
  message = FALSE
)
```

# inflationinequality

<!-- badges: start -->
[![R-CMD-check](https://github.com/jeremimontornes/inflationinequality/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/jeremimontornes/inflationinequality/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

`inflationinequality` provides methods to calculate and visualize inflation inequality indicators.

Package version: `0.0.2`

## Features

- Calculate and visualize inflation and contributions to inflation by households categories
- Calculate inflation burden as a share of disposable income
- Simulate counterfactual price indices

## Methodological workflow

The main workflow is:

`load_*()` -> `calculate_weights()` -> `calculate_price_indices()` / `calculate_inflation()` -> `calculate_contributions()` -> plots / validation.

For aggregation-bias checks, the package uses:

`aggregation_bias = inflation_gap_coicop_upperdigit - inflation_gap_coicop_lowerdigit`

where `upperdigit` is the more aggregated COICOP calculation and `lowerdigit` is the more detailed COICOP calculation.

## R workflow structure

```text
R/load_data.R                         # Download HICP prices, HICP weights, and Eurostat HBS data
├── load_cpi()                        # Monthly HICP price indices
├── load_index_weights()              # Annual HICP item weights
└── load_hbs()                        # HBS expenditure shares by household category

R/calculate_weights.R                 # Match HICP weights with HBS shares
└── calculate_weights()               # Category-specific COICOP weights

R/calculate_contributions.R           # Compute COICOP contributions to inflation
└── calculate_contributions()

R/calculate_inflation.R               # Aggregate contributions into inflation indicators
├── calculate_inflation()
├── calculate_total_inflation()
└── calculate_inflation_gap()

R/calculate_price_indices.R           # Build chained HICP price indices in level
└── calculate_price_indices()         # Uses INSEE HBS level 3 automatically for France income level 3

R/calculate_inflation_burden.R        # Inflation cost burden by household group
├── load_consumption_to_income()
├── load_consumption_expenditure()
└── calculate_inflation_burden()

R/compare_to_official_hicp.R          # Validation against published all-items HICP
└── compare_to_official_hicp()

R/plot_inflation.R                    # Standard plots
├── plot_time_series()
├── plot_grouped_bar()
├── plot_weight_shares()
├── plot_group_price_indices()
├── plot_inflation_burden()
├── plot_contribution_gap()
└── plot_inflation_gap()

R/simulate.R                          # Counterfactual CPI scenarios
└── simulate_cpi()

R/coicop_bridge.R                     # COICOP matching and bridge tables
├── build_coicop_bridge()
└── write_coicop_bridge_html()

```

COICOP bridge HTML tables:

- [France HICP-HBS bridge, income level 2](https://jeremimontornes.github.io/inflationinequality/france_coicop_hicp_hbs_bridge_income_level2_2010_2026.html)
- [France HICP-HBS bridge audit, income level 2](https://jeremimontornes.github.io/inflationinequality/france_coicop_hicp_hbs_bridge_income_level2_2010_2026_audit.html)
- [France HICP-HBS bridge, INSEE income level 3](https://jeremimontornes.github.io/inflationinequality/france_coicop_hicp_hbs_bridge_income_level3_insee_2010_2026.html)
- [France HICP-HBS bridge audit, INSEE income level 3](https://jeremimontornes.github.io/inflationinequality/france_coicop_hicp_hbs_bridge_income_level3_insee_2010_2026_audit.html)
- [France ECOICOP v2 to COICOP v1 level 3 audit](https://jeremimontornes.github.io/inflationinequality/france_hicp_ecoicopv2_to_coicopv1_level3_audit.html)

## Example

Let us visualize inflation inequality across income quintiles in France since 2019, using the France level 3 COICOP workflow.

```{r example}
library(inflationinequality)
inflation <- calculate_inflation("FR", "income", start_year = 2019, level = 3)
plot_time_series(inflation)
```

Source: Eurostat and national statistical institutes, HICP-HBS. Note: year-on-year change

## Methodological metadata

The table below summarises the main methodological choices currently made by the package and identifies which choices users can override. It is intended as a living reference for handling country-specific cases.

```{r methodology-metadata, echo = FALSE, results = "asis"}
load("data/methodology_metadata.rda")
cat("<small>\n")
methodology_metadata |>
  dplyr::select(topic, package_default, user_parameter, user_can_change, notes) |>
  knitr::kable(format = "html", row.names = FALSE) |>
  print()
cat("\n</small>\n")
```

### Country-specific COICOP data caveats

```{r country-coicop-special-cases, echo = FALSE, results = "asis"}
load("data/methodology_metadata.rda")
load("data/country_coicop_special_cases.rda")
cat("<small>\n")
country_coicop_special_cases |>
  dplyr::select(country, coicop_case, population_group_caveat, housing_caveat, package_status) |>
  knitr::kable(format = "html", row.names = FALSE) |>
  print()
cat("\n</small>\n")
```

## Vignettes

The introduction vignette walks through the package methodology, core functions, and plotting helpers:

- [Introduction to inflationinequality](https://jeremimontornes.github.io/inflationinequality/articles/inflationinequality-intro.html)
- [Advanced options](https://jeremimontornes.github.io/inflationinequality/articles/advanced-options.html)
- [Using national data](https://jeremimontornes.github.io/inflationinequality/articles/using-custom-data.html)
- [Calculating inflation burden](https://jeremimontornes.github.io/inflationinequality/articles/inflation-burden.html)
- [Verifying calculated inflation](https://jeremimontornes.github.io/inflationinequality/articles/verifying-calculated-inflation.html)
