---
output:
  github_document:
    pandoc_args: ["--wrap=none", "--from=markdown-smart"]
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  dev = "svg",
  warning = FALSE,
  message = FALSE
)
```

# inflationinequality

<!-- badges: start -->
[![R-CMD-check](https://github.com/jeremimontornes/inflationinequality/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/jeremimontornes/inflationinequality/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

`inflationinequality` provides methods to calculate and visualize inflation inequality indicators.

## Features

- Calculate and visualize inflation and contributions to inflation by households categories
- Simulate counterfactual price indices

## Vignettes

The introduction vignette walks through the package methodology, core functions, and plotting helpers:

- [Introduction to inflationinequality](vignettes/inflationinequality-intro.Rmd)
- [Online introduction page](https://jeremimontornes.github.io/inflationinequality/)
- [Versioning and release rules](VERSIONING.md)

After installing the package with vignettes, open it in R with:

```{r, eval = FALSE}
vignette("inflationinequality-intro", package = "inflationinequality")
```

## R workflow structure

```text
R/load_data.R                         # Download HICP prices, HICP weights, and Eurostat HBS data
├── load_cpi()                        # Monthly HICP price indices
├── load_index_weights()              # Annual HICP item weights
└── load_hbs()                        # HBS expenditure shares by household category

R/calculate_weights.R                 # Match HICP weights with HBS shares
└── calculate_weights()               # Category-specific COICOP weights

R/calculate_contributions.R           # Compute COICOP contributions to inflation
└── calculate_contributions()

R/calculate_inflation.R               # Aggregate contributions into inflation indicators
├── calculate_inflation()
├── calculate_total_inflation()
└── calculate_inflation_gap()

R/calculate_price_indices.R           # Build chained HICP price indices in level
└── calculate_price_indices()         # Uses INSEE HBS level 3 automatically for France income level 3

R/compare_to_official_hicp.R          # Validation against published all-items HICP
└── compare_to_official_hicp()

R/plot_inflation.R                    # Standard plots
├── plot_time_series()
├── plot_grouped_bar()
├── plot_contribution_gap()
└── plot_inflation_gap()

R/simulate.R                          # Counterfactual CPI scenarios
└── simulate_cpi()

R/coicop_bridge.R                     # COICOP matching and bridge tables
├── build_coicop_bridge()
└── write_coicop_bridge_html()

scripts/plot_fr_hicp_validation.R     # France validation plots: calculated vs published HICP
scripts/plot_fr_q1_q5_base2010_to_2026_03.R
                                      # France Q1/Q5 index-level plot
scripts/build_fr_coicop_bridge_html.R # France HICP-HBS bridge HTML tables
scripts/build_fr_ecoicopv2_to_coicopv1_level3_audit_html.R
                                      # France ECOICOP v2 -> COICOP v1 level 3 audit HTML
```

## Example

Let us visualize inflation inequality across income quintiles in Italy since 2019.

```{r example}
library(inflationinequality)
inflation <- calculate_inflation("IT", "income", start_year = 2019)
plot_time_series(inflation)
```

Source: Eurostat and national statistical institutes, HICP-HBS. Note: year-on-year change

## Methodological metadata

The table below summarises the main methodological choices currently made by the package and identifies which choices users can override. It is intended as a living reference for handling country-specific cases.

```{r methodology-metadata, echo = FALSE, results = "asis"}
load("data/methodology_metadata.rda")
cat("<small>\n")
methodology_metadata |>
  dplyr::select(topic, package_default, user_parameter, user_can_change, notes) |>
  knitr::kable(format = "html", row.names = FALSE) |>
  print()
cat("\n</small>\n")
```
