---
output:
  github_document:
    pandoc_args: ["--wrap=none", "--from=markdown-smart"]
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  dev = "svg",
  warning = FALSE,
  message = FALSE
)
```

# inflationinequality

<!-- badges: start -->
[![R-CMD-check](https://github.com/jeremimontornes/inflationinequality/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/jeremimontornes/inflationinequality/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

`inflationinequality` provides methods to calculate and visualize inflation inequality indicators.

## Features

- Calculate and visualize inflation and contributions to inflation by households categories
- Simulate counterfactual price indices

## Vignettes

The introduction vignette walks through the package methodology, core functions, and plotting helpers:

- [Introduction to inflationinequality](vignettes/inflationinequality-intro.Rmd)
- [Online introduction page](https://jeremimontornes.github.io/inflationinequality/)

After installing the package with vignettes, open it in R with:

```{r, eval = FALSE}
vignette("inflationinequality-intro", package = "inflationinequality")
```

## Example

Let us visualize inflation inequality across income quintiles in Italy since 2019.

```{r example}
library(inflationinequality)
inflation <- calculate_inflation("IT", "income", start_year = 2019)
plot_time_series(inflation)
```

## Methodological metadata

The table below summarises the main methodological choices currently made by the package and identifies which choices users can override. It is intended as a living reference for handling country-specific cases.

```{r methodology-metadata, echo = FALSE}
load("data/methodology_metadata.rda")
methodology_metadata |>
  dplyr::select(topic, package_default, user_parameter, user_can_change, notes) |>
  knitr::kable(row.names = FALSE)
```
