---
title: "Advanced options"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced options}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  warning = FALSE,
  message = FALSE
)
```

```{r setup, eval = TRUE}
library(inflationinequality)
options(rdbnomics.use_readLines = TRUE)
```

This vignette demonstrates the more advanced features in `inflationinequality`.

***

## Ensure complete CPI

It's possible that in some CPI datasets, price data of certain product categories are unavailable in certain time periods. For example, medical consultation services or "062" are usually missing price data in early years.

In order to fill these gaps, we can synthesize the price data using the price movements of its parent product category. In the case of the "062", its parent is health services or "06", so we synthesize the missing price data of "062" using the price movements of "06".

Let's first load all CPI price data in France.

```{r}
cpi_fr <- load_cpi("FR", level = 3)
```

We can see that there are some products missing price data in 1996:

```{r}
setdiff(cpi_fr$dt[, unique(coicop)], cpi_fr$dt[year == 1996, unique(coicop)])
```

We can also use the `get_missing_cpi_tuples()` function to retrieve all the missing data points.

```{r}
get_missing_cpi_tuples(cpi_fr)
```

To correct the CPI object, we simply use the `correct_cpi()` method.

```{r}
corrected_cpi_fr <- correct_cpi(cpi_fr)
```

As a result, there are no missing data points:

```{r}
get_missing_cpi_tuples(corrected_cpi_fr)
```

As a matter of convenience, this functionality has been added as an option in `calculate_contributions()` and `calculate_inflation()`.

```{r, eval = FALSE}
inflation <- calculate_inflation("FR", "income", level = 3, ensure_complete_cpi = TRUE)
```

### Using sparse data

When the data is limited, there may be errors when synthesizing the price data:

```{r}
cpi_limited <- load_cpi("FR", level = 3, start_year = 1996, end_year = 2000)
corrected_cpi_limited <- correct_cpi(cpi_limited)
```

## Linear interpolation in HBS

Household Budget Surveys (HBS) are typically published in waves every five years.
The change in weights between one HBS and the next can be drastic, which can unrealistically impact the measured inflation rate.
To address this, we propose linearly interpolating the HBS weights to smooth out the changes in weights over time.

Here is an example of a HBS dataset:

```{r, eval = FALSE}
hbs <- load_hbs("FR", "income", start_year = 2005)
```

We can apply linear interpolation by using `interpolate_hbs()` on the `"hbs"` object.

```{r, eval = FALSE}
interpolated_hbs <- interpolate_hbs(hbs)
```

For convenience, we can use the option `interpolated_hbs` in `calculate_contributions()` or `calculate_inflation()`.

```{r, eval = FALSE}
interpolated_inflation <- calculate_inflation("FR", "income", interpolated_hbs = TRUE)
```

## Specific HBS years

We can choose to apply a specific year's HBS weights for all months when calculating inflation rates (or contributions to inflation).

```{r, eval = FALSE}
calculate_inflation("FR", "income", specific_hbs_year = 2005)
```

Alternatively, we can linearly interpolate the HBS weights so we can choose any year's HBS weights.

```{r, eval = FALSE}
calculate_inflation("FR", "income", interpolated_hbs = TRUE, specific_hbs_year = 2019)
```

## Aggregation-bias table

Aggregation bias can be checked by comparing the annualized inflation gap
between the bottom and top income groups at two COICOP levels. Here, "upper"
means the more aggregated COICOP level, with fewer digits, and "lower" means the
more detailed COICOP level, with more digits. Let $g_{Q1}^{(U)}$ and
$g_{Q5}^{(U)}$ denote the annualized growth rates of the chained Laspeyres price
indices for the first and fifth quintiles computed at the upper COICOP level
$U$. Let $g_{Q1}^{(L)}$ and $g_{Q5}^{(L)}$ denote the same growth rates computed
at the lower COICOP level $L$. The reported gaps are:

$$
inflation\_gap\_coicop\_upperdigit = g_{Q1}^{(U)} - g_{Q5}^{(U)}
$$

$$
inflation\_gap\_coicop\_lowerdigit = g_{Q1}^{(L)} - g_{Q5}^{(L)}
$$

The output keeps the shorter column names `coarse_gap` for
`inflation_gap_coicop_upperdigit` and `fine_gap` for
`inflation_gap_coicop_lowerdigit`. The aggregation-bias measure is:

$$
aggregation\_bias = inflation\_gap\_coicop\_upperdigit - inflation\_gap\_coicop\_lowerdigit
$$

The package's default `formula = "laspeyres"` preserves the Eurostat-style
chained Laspeyres convention: it aggregates monthly unchained price relatives
with the available annual category weights and then chains those monthly
aggregates over time. This is why the substitution-bias table uses
`chained_laspeyres_growth` as the reference side.

The comparison below applies this calculation to six countries between
`level = 1` and `level = 2`, using the same time convention as the
substitution-bias example, starting in 2019:

```{r}
data.table::rbindlist(
  lapply(
    c("DE", "FR", "IT", "ES", "NL", "BE"),
    function(country) {
      calculate_aggregation_bias(
        country,
        "income",
        coarse_level = 1,
        fine_level = 2,
        start_year = 2019
      )
    }
  )
)
```

Example output:

| country | category_type | coarse_level | fine_level | coarse_gap | fine_gap | aggregation_bias |
|:---|:---|---:|---:|---:|---:|---:|
| DE | income | 1 | 2 | 0.112 | -0.503 | 0.615 |
| FR | income | 1 | 2 | -0.083 | -0.158 | 0.075 |
| IT | income | 1 | 2 | 0.000 | -0.074 | 0.074 |
| ES | income | 1 | 2 | 0.169 | -0.285 | 0.454 |
| NL | income | 1 | 2 | 0.084 | -0.011 | 0.095 |
| BE | income | 1 | 2 | 0.016 | 0.056 | -0.040 |

If a country has an integrated source at `level = 3`, `calculate_aggregation_bias()`
uses that finer level by default and compares it with the previous level. France
income calculations are currently the implemented case: by default, the function
compares `level = 2` with the bundled INSEE `level = 3` workflow.

## Substitution-bias table

The BLS compares a fixed-basket research CPI (`R-CPI-I`) with a chained Tornqvist-formula research CPI (`R-C-CPI-I`). For the HICP setting, the package keeps the Eurostat-style chained Laspeyres convention on the reference side. The substitution-bias measure is therefore:

$$
substitution\_bias = chained\_laspeyres\_growth - chained\_toernqvist\_growth
$$

The same idea can be applied to France with Eurostat HICP and HBS data:

```{r}
calculate_substitution_bias(
  "FR",
  "income",
  level = 2,
  start_year = 2019
)
```

Example output:

| category | chained_laspeyres_growth | chained_toernqvist_growth | substitution_bias |
|:---|---:|---:|---:|
| Total | 2.96 | 2.90 | 0.06 |
| First quintile | 2.83 | 2.74 | 0.10 |
| Second quintile | 2.94 | 2.85 | 0.10 |
| Third quintile | 2.98 | 2.90 | 0.09 |
| Fourth quintile | 3.01 | 2.93 | 0.08 |
| Fifth quintile | 2.99 | 2.92 | 0.07 |

The returned columns are:

- `chained_laspeyres_growth`: annualized growth rate of the package's Eurostat-style chained Laspeyres aggregation.
- `chained_toernqvist_growth`: annualized growth rate of a chained Tornqvist aggregation computed with `hicp::toernqvist()`.
- `substitution_bias`: `chained_laspeyres_growth - chained_toernqvist_growth`, in percentage points.

This is not a perfect replication of the BLS table, because Eurostat HBS income-quintile expenditure shares are not available at the same monthly frequency as the U.S. Consumer Expenditure Survey inputs used by BLS. The package therefore uses annual category-specific weights and approximates the Tornqvist comparison with previous-year and current-year weights, while retaining the HICP chained-Laspeyres convention for the reference index.

