---
title: "Introduction to inflationinequality"
output: rmarkdown::html_document
vignette: >
  %\VignetteIndexEntry{Introduction to inflationinequality}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

```{r setup, include = FALSE}
library(inflationinequality)
options(rdbnomics.use_readLines = TRUE)
```

This vignette introduces the fundamental features of the `inflationinequality` package: how to compute inflation and contributions to inflation across demographics, how to download CPI and HBS datasets, and useful plotting functions.

***

## What is the purpose of this package?

Inflation is the general and prolonged increase of prices. Different demographic categories have different spending habits and as a result, some are more affected than inflation by others. This package provides methods that calculates and visualizes inflation and contribution to inflation by demographic category, using public data. Currently, the package only officially supports Euro area (EA-19) countries.

## How is inflation calculated?

Let's say for example we want to compare inflation rates for different income categories.

In order to calculate inflation, we use Household Budget Surveys (HBS) to calculate the percentage of consumption that each level 1 or level 2 COICOP category makes up for all income quintiles.

NOTE: Eurostat harmonizes HBS income data by quintiles, but we can include HBS data with different categories like deciles in the case of INSEE (France).
NOTE: In keeping with the Eurostat HICP methodology, we do not include imputed rent (CP042) in our analysis.

### Weights formula

We first calculate the ratio between the expenditure for each income quintile of this product $j$ and that observed for all households as defined by the HBS at year $y'$, 
denoted 

$$
\text{HBS}^{y'}_{j, \text{quintile}}
$$ 

and

$$
\text{HBS}^{y'}_{j, \text{total}}
$$ 

respectively.

This term "corrects" the annual CPI weight corresponding to product $j$, denoted $w^y_{\text{CPI},j}$, therefore

$$
w^y_{j, \text{quintile}} = w^y_{\text{CPI},j} \times \text{HBS}^{y'}_{j, \text{quintile}} / \text{HBS}^{y'}_{j, \text{total}}
$$

After this correction, the weights are then renormalized to sum to 100 for each income quintile.

You may notice that the year of the HBS data $y'$ and the year of the CPI weight $y$ are different. HBS data are released every five years while the CPI is released monthly so the following rules are established to properly align the HBS data with the CPI:

1. For a given CPI weight released in year $y$, we use the most recent HBS data $y'$.
2. If there is no HBS data at the same year or earlier than $y$, then we use the earliest HBS data available.

These operations are implemented in `calculate_weights()` and data is downloaded using `load_index_weights()` and `load_hbs()`.

### Contributions to inflation formula

The CPI is chained index whose weightings change each year. We calculate the contributions using the following formula:

$$
\text{Contrib}^{y,m}_{j,\text{quintile}} = \frac{p^{y-1,12}}{p^{y-1,m}}\times w^{y-1}_{j,\text{quintile}} \times \bigg(\frac{p^{y,m}_j - p^{y-1,12}_j}{p^{y-1,12}_j} \bigg) + \frac{p^{y-2,12}}{p^{y-1,m}} \times w^{y-2}_{j,\text{quintile}} \times \bigg(\frac{p^{y-1,12}_j - p^{y-1,m}_j}{p^{y-2,12}_j} \bigg)
$$

where

* $p^{y,m}_j$ is the price index of product $j$ in month $m$ of year $y$
* $p^{y,m}$ is the price index of the product basket in month $m$ of year $y$

The formula is implemented in `calculate_contributions()` which downloads the price index 
* $p^{y,m}_j$
 using `load_cpi()` and calculates the weights 
* $w^{y}_{j,\text{quintile}}$ with `calculate_weights()`.   

### Inflation and inflation gap formula

Inflation in a given month for each quintile results from the sum of the contributions of product $j$.

$$
\text{Inflation}^{y,m}_\text{quintile} = \sum_j \text{Contrib}^{y,m}_{j,\text{quintile}}
$$

This formula is implemented in `calculate_inflation()`.

By calculating the inflation by income quintile, we can calculate the inflation gap to measure inflation equalities:

$$
InflationGap^{y,m} = Inflation^{y,m}_{quintile1} - Inflation^{y,m}_{quintile5}
$$

This formula is implemented in `calculate_inflation_gap()`.

## Calculating inflation using `inflationinequality`

In `inflationinequality`, the function `calculate_inflation()` computes the inflation rate of a given country and across a specified demographic category. For example, we can calculate French inflation rates from 2019 onwards by income category: 

```{r}
inflation_fr <- calculate_inflation("FR", "income", start_year = 2019)
inflation_fr$dt
```

Notice that `inflation_fr` is an object of class `"inflation"` which is composed of the actual inflation data in the component `dt` but is also composed of useful metadata: `categories` (an ordered list of demographic classifications), the `start_year` and `start_month`, and the `end_year` and `end_month`.

The full documentation for the `"inflation"` object can be viewed by launching `?calculate_inflation`.

The `inflationinequality` package uses objects to effectively handle the metadata associated with different datasets.

We can access the component `dt` using the `$` operator:

```{r}
head(inflation_fr$dt)
```

In the inflation dataset, the annualized inflation rate has been computed for every month since 2019 and for each income category.

We can visualize this time series with `ggplot2`.

```{r}
library(ggplot2)

ggplot(inflation_fr$dt, aes(x = as.Date(ISOdate(year, month, 1)), y = inflation, color = factor(category, levels = inflation_fr$categories), group = category)) +
  geom_line() +
  scale_x_date(date_labels = "%Y", date_breaks = "1 year") +
  labs(
    title = "French inflation by income category",
    x = "Date",
    y = "Annualized inflation rate"
  ) +
  scale_color_manual(
    name = "Income category",
    values = c("red", "darkorange", "yellow", "green", "darkgreen"),
    breaks = inflation_fr$categories
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

We can also calculate the contributions to inflation. Here, we're calculating contributions to inflation in Spain by income quintiles from 2019 to 2023.

```{r}
contributions_es <- calculate_contributions("ES", "income", start_year = 2019, end_year = 2023)
head(contributions_es$dt)
```

## Plotting graphs: time series, grouped bar charts, and more

The `inflationinequality` package also provides useful plotting functions.

### Time series

Using the French inflation rates that we just calculated, we can plot the inflation rates as a time series according to their demographic. Here, the `plot_time_series()` function allows us to compare how inflation disproportionately affects the top and bottom income quintiles (or categories, in general) and also displays the average inflation over the entire population.

```{r}
plot_time_series(inflation_fr)
```

We can also visualize the inflation regime in Italy since 2019.

```{r}
inflation_it <- calculate_inflation("IT", "income", start_year = 2019)
plot_time_series(inflation_it)
```

### Grouped bar charts

We can also plot a grouped bar chart to compare inflation in different countries for a specific year. In this case, we are comparing 2022 inflation rates in Germany and Italy by income quintile.

```{r}
inflation_de <- calculate_inflation("DE", "income", start_year = 2022, end_year = 2022)
plot_grouped_bar(2022, inflation_de, inflation_it)
```

### Contributions to inflation gap chart

Furthermore, we can plot the contributions to the inflation gap with `plot_contribution_gap()`.

```{r}
plot_contribution_gap(contributions_es)
```

The above selected product categories in the legend are the default selection, but you can customize the selection by setting the `category_mapping` parameter as follows:

```{r}
plot_contribution_gap(
  contributions_es,
  category_mapping = list(
    "Food" = "011",
    "Water, electricity, gas and other fuels" = c("044", "045")
  )
)
```

## Loading CPI data or HBS data

Additionally, you can download the CPI and HBS datasets.

Let's say we want to download all CPI data from France.

```{r}
cpi <- load_cpi("FR")
head(cpi$dt)
```

You can also choose the month since CPI data is released monthly:

```{r}
cpi <- load_cpi("IT", start_year = 2019, start_month = 5, end_year = 2020, end_month = 3)
head(cpi$dt)
```

By default, the package will download the dataset the highest granularity of COICOP levels, which in this case is level 2. We can also choose to download only COICOP level 1.

```{r}
cpi <- load_cpi("ES", level = 1)
head(cpi$dt)
```

Note that `load_cpi()` does not include the index weights.

If you want to download the CPI weights, then use `index_weights()`, for example

```{r}
index_weights <- load_index_weights("ES")
head(index_weights$dt)
```

You can also download HBS data by demographic category. Let's download all available French HBS data by income.

```{r}
hbs <- load_hbs("FR", "income")
head(hbs$dt)
```

Note that HBS data is released periodically, not annually, so downloading within a short timeframe may result in no data being downloaded at all.

```{r}
hbs <- load_hbs("FR", "age", start_year = 2016, end_year = 2017)
head(hbs$dt)
```
