---
title: "Simulating counterfactual indices"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulating counterfactual indices}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "svg",
  warning = FALSE,
  message = FALSE
)
```

```{r setup}
library(inflationinequality)
options(rdbnomics.use_readLines = TRUE)
```

The package also provides a method to simulate counterfactual indices by synthesizing CPI price data according to specific parameters.

***

In 2022, European countries experienced high inflation rates, especially from electricity and natural gas prices.
Almost all European governments implemented measures to reduce the inflationary forces with price caps or subsidies.
In this example, we will simulate a counterfactual index for fuel between April 2022 and February 2023, with the hypothesis that without government intervention, the price index for petrol would have increased by 40% on April 2022 and remained the same throughout the inflation period.

Let's first download the CPI dataset. Below is the actual inflation rate.

```{r}
INSEE_HBS_2017 <- readRDS("articles/INSEE_HBS_2017_level3.RDS")
cpi <- load_cpi("FR", level = 3)
inflation <- calculate_inflation("FR", "income", level = 3, start_year = 2019, custom_cpi = cpi, custom_hbs = INSEE_HBS_2017)
plot_time_series(inflation)
```

We can measure the contribution of fuel price to the inflation rate better with `plot_contribution_gap()`.

```{r}
contributions <- calculate_contributions("FR", "income", level = 3, start_year = 2019, custom_cpi = cpi, custom_hbs = INSEE_HBS_2017)
plot_contribution_gap(contributions, category_mapping = list("Fuel" = "0722"))
```

Now, we can simulate the shock with the associated shock parameters in this data frame.

```{r}
simulations <- data.frame(
  coicop = "0722",
  shock = 40,
  start_year = 2022,
  start_month = 4,
  end_year = 2023,
  end_month = 2
)
sim_cpi <- simulate_cpi(cpi, simulations)
sim_inflation <- calculate_inflation("FR", "income", level = 3, start_year = 2019, custom_cpi = sim_cpi, custom_hbs = INSEE_HBS_2017)
plot_time_series(sim_inflation)
```

And using the contribution gap plot:

```{r}
sim_contributions <- calculate_contributions("FR", "income", level = 3, start_year = 2019, custom_cpi = sim_cpi, custom_hbs = INSEE_HBS_2017)
plot_contribution_gap(sim_contributions, category_mapping = list("Fuel" = "0722"))
```
