---
title: "Simulating counterfactual indices"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulating counterfactual indices}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  warning = FALSE,
  message = FALSE
)
```

```{r setup, eval = TRUE}
library(inflationinequality)
```

The package also provides a method to simulate counterfactual indices by synthesizing CPI price data according to specific parameters.

***

In 2022, European countries experienced high inflation rates, especially from electricity and natural gas prices.
Almost all European governments implemented measures to reduce the inflationary forces with price caps or subsidies.
In this example, we will simulate a counterfactual index for fuel between April 2022 and February 2023, with the hypothesis that without government intervention, the price index for petrol would have increased by 40% on April 2022 and remained the same throughout the inflation period.

Let's first download the CPI dataset. Below is the actual inflation rate.

```{r}
INSEE_HBS_2017 <- readRDS("articles/INSEE_HBS_2017_level3.RDS")
cpi <- load_cpi("FR", level = 3)
inflation <- calculate_inflation("FR", "income", level = 3, start_year = 2019, custom_cpi = cpi, custom_hbs = INSEE_HBS_2017)
plot_time_series(inflation)
```

We can measure the contribution of fuel price to the inflation rate better with `plot_contribution_gap()`.

```{r}
contributions <- calculate_contributions("FR", "income", level = 3, start_year = 2019, custom_cpi = cpi, custom_hbs = INSEE_HBS_2017)
plot_contribution_gap(contributions, category_mapping = list("Fuel" = "0722"))
```

Now, we can simulate the shock with the associated shock parameters in this data frame.

```{r}
simulations <- data.frame(
  coicop = "0722",
  shock = 40,
  start_year = 2022,
  start_month = 4,
  end_year = 2023,
  end_month = 2
)
sim_cpi <- simulate_cpi(cpi, simulations)
sim_inflation <- calculate_inflation("FR", "income", level = 3, start_year = 2019, custom_cpi = sim_cpi, custom_hbs = INSEE_HBS_2017)
plot_time_series(sim_inflation)
```

And using the contribution gap plot:

```{r}
sim_contributions <- calculate_contributions("FR", "income", level = 3, start_year = 2019, custom_cpi = sim_cpi, custom_hbs = INSEE_HBS_2017)
plot_contribution_gap(sim_contributions, category_mapping = list("Fuel" = "0722"))
```

## Simulating several policy counterfactuals

For policy scenarios, it is often useful to define each measure separately and
then apply all measures to the CPI object in one step. This avoids recalculating
the aggregate CPI basket after every individual measure.

The helper `cpi_counterfactual_policy()` creates one monthly policy table. The
function `simulate_cpi_policy_counterfactuals()` can then apply several of these
tables simultaneously and recalculate the aggregate CPI basket once.

The example below defines five VAT counterfactuals. Each row-binding step adds
one product-specific measure to the scenario.

```{r}
vat_policies <- data.table::rbindlist(list(
  cpi_counterfactual_policy(
    policy_id = "vat_bread",
    coicop = "0111",
    type = "vat",
    start = "2022-04",
    end = "2022-12",
    old_rate = 0.20,
    new_rate = 0.055
  ),
  cpi_counterfactual_policy(
    policy_id = "vat_meat",
    coicop = "0112",
    type = "vat",
    start = "2022-04",
    end = "2022-12",
    old_rate = 0.20,
    new_rate = 0.055
  ),
  # [...]
))

policy_cpi <- simulate_cpi_policy_counterfactuals(
  cpi,
  policies = vat_policies,
  recalculate_price_basket = TRUE
)

policy_inflation <- calculate_inflation(
  "FR",
  "income",
  level = 3,
  start_year = 2019,
  custom_cpi = policy_cpi,
  custom_hbs = INSEE_HBS_2017
)
plot_time_series(policy_inflation)
```

The aggregate CPI basket is stored in `dt_basket`. The example below changes
only the simulated sub-indices, keeps all other French CPI items unchanged, and
then compares the observed and counterfactual total year-on-year inflation
rates.

```{r total-cpi-counterfactual-plot, eval = FALSE, fig.width = 7, fig.height = 4}
cpi_fr <- load_cpi("FR", level = 3, start_year = 2019, end_year = 2026)

policy_cpi_fr <- simulate_cpi_policy_counterfactuals(
  cpi_fr,
  policies = data.table::rbindlist(list(
    cpi_counterfactual_policy(
      policy_id = "electricity_vat_counterfactual",
      coicop = "0451",
      type = "vat",
      start = "2022-04",
      end = "2023-12",
      old_rate = 0.20,
      new_rate = 0.055
    ),
    cpi_counterfactual_policy(
      policy_id = "gas_tariff_counterfactual",
      coicop = "0452",
      type = "ratio",
      start = "2022-04",
      end = "2023-12",
      ratio = 1.25
    ),
    cpi_counterfactual_policy(
      policy_id = "fuel_subsidy_counterfactual",
      coicop = "0722",
      type = "unit_subsidy",
      start = "2022-04",
      end = "2022-12",
      subsidy = 0.18,
      unit_price = 1.85
    )
  )),
  recalculate_price_basket = TRUE
)

total_indices <- data.table::rbindlist(list(
  data.table::copy(cpi_fr$dt_basket)[, scenario := "Observed inflation"],
  data.table::copy(policy_cpi_fr$dt_basket)[, scenario := "Counterfactual inflation"]
))
total_indices[, date := as.Date(sprintf("%04d-%02d-01", year, month))]
data.table::setorder(total_indices, scenario, date)
total_indices[, inflation_yoy := 100 * (value / data.table::shift(value, 12) - 1), by = scenario]
total_indices <- total_indices[
  date >= as.Date("2020-01-01") &
    date <= min(as.Date("2026-12-01"), max(date, na.rm = TRUE)) &
    !is.na(inflation_yoy)
]

ggplot2::ggplot(
  total_indices,
  ggplot2::aes(x = date, y = inflation_yoy, color = scenario)
) +
  ggplot2::geom_hline(yintercept = 0, linewidth = 0.3, color = "grey75") +
  ggplot2::geom_line(linewidth = 0.8) +
  ggplot2::labs(x = NULL, y = "Year-on-year inflation (%)", color = NULL) +
  ggplot2::theme_minimal()
```

The same workflow can mix different policy types. For example, a VAT
counterfactual can be combined with a unit fuel subsidy or an externally
supplied tariff index before recalculating the CPI basket.
