---
title: "Using national CPI-HBS data"
output: rmarkdown::html_document
vignette: >
  %\VignetteIndexEntry{Using national CPI-HBS data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

```{r setup, include = FALSE}
library(inflationinequality)
options(rdbnomics.use_readLines = TRUE)
```

This vignette introduces methods to include your own data.

***

## Downloading data from Eurostat

Using DBnomics, we download CPI and HBS data from any country in the Eurozone and compare with those of another, due to having standardized definitions of categories, etc.

However, Eurostat  only provides 2-digit COICOP HBS data. For some countries (France, Italy, etc.), instead of using the data published by Eurostat, data published directly by national statistical institutes can be used like France's INSEE or Italy's Istat.

In this vignette, we will download datasets in CSV format and format the data for the package.

## Downloading CPI from INSEE

Let's download CPI data from INSEE, sourced from this page: [https://www.insee.fr/fr/statistiques/series/102342213](https://www.insee.fr/fr/statistiques/series/102342213). In particular, this CPI dataset is bundled with other data that we do not need so we'll have to process this data.

First, we load some necessary libraries.

```{r}
# Required for fast CSV reading
library(readr)

# Required for reading Excel spreadsheets
library(readxl)

# Required for manipulating data.table objects,
# which is the primary data format for this package
library(data.table)

# Required for the %>% pipe operator
library(dplyr)

# Required for some string operations
library(stringr)

# Required for some date operations
library(lubridate)
```

We'll use the CSV files since they're easier to process into an R object.
Download the ZIP archive for the CSV files from the website and extract the archive.

We'll first load `valeurs_mensuelles.csv` (monthly values) into a data.table.

```{r}
# Get CSV file path
monthly_values_csv_file <- "famille_IPC-2015/valeurs_mensuelles.csv"

# Read CSV data into a data.table
INSEE_CPI_monthly_values <- as.data.table(read_csv2(monthly_values_csv_file))
```

The following section will be dedicated to correctly formatting this dataset, which is essential for loading data from sources other than Eurostat to the `inflationinequality` package.

### Correctly formatting a time series dataset

#### Formatting CPI per product category

This dataset bundles together monthly CPI index data with other data like year-on-year change, which we do not need.
In addition, the COICOP codes are written in the series name (identifier) or "Libellé" so we'll have extract that.
The dataset is structured in a *wide* format, rather than the *long* format preferred by our package.
In wide format, each date is assigned its own column, excluding other identifiers such as the COICOP code.
This structure is less conducive to efficient data analysis compared to the long format.
We will write the code will clean the dataset into a format that is readable for our package.

We'll also filter for level 2 COICOP codes, that is 3-digit COICOP.

First, we want to remove the columns `idBank`, `Dernière mise à jour` and `Période` because they're useless.

```{r}
INSEE_CPI_dt <- INSEE_CPI_monthly_values %>%
  .[, `:=`(
    idBank = NULL,
    `Dernière mise à jour` = NULL,
    `Période` = NULL)]
```

Next, we use `str_extract()` to find the COICOP code in the `Libellé` column.
The regex pattern looks for:
- A positive lookbehind for the specific text preceding the code
- Two digits followed by up to three occurrences of a dot and one or more digits

```{r}
INSEE_CPI_dt <- INSEE_CPI_dt %>%
  .[, coicop := str_extract(
    `Libellé`,
    "(?<=Indice des prix à la consommation - Base 2015 - Ensemble des ménages - France - Nomenclature Coicop : )\\d{2}(\\.\\d+){0,3}"
  )]
```

We also need to remove all NA values in the `coicop` column, in case there were no matches in `Libellé`.

```{r}
INSEE_CPI_dt <- INSEE_CPI_dt %>%
  .[!is.na(coicop), ]
```

Then we use `str_remove_all()` to remove all dots from the extracted COICOP code.
The regex pattern `"\\."` matches any dot character.

```{r}
INSEE_CPI_dt <- INSEE_CPI_dt %>%
  .[, coicop := str_remove_all(coicop, "\\.")]
```

Now We'll reshape data from wide to long format.

```{r}
INSEE_CPI_dt <- INSEE_CPI_dt %>%
  melt(id.vars = c("Libellé", "coicop"), 
                   variable.name = "date", value.name = "value")
```


Finally, we select and format necessary columns.
Remember, in a "cpi" object, the "dt" component must (in any order) have the columns: `coicop`, `series_name`, `year`, `month`, `value`

```{r}
INSEE_CPI_dt <- INSEE_CPI_dt %>%
  .[, .(coicop,
    series_name = `Libellé`,
    year = year(ym(as.character(date))),
    month = month(ym(as.character(date))),
    value = as.numeric(value)
  )]
```

Let's see the layout of our newly formatted dataset.

```{r}
# Ignore the series_name column because it is very long
head(INSEE_CPI_dt[, .SD, .SDcols = !c("series_name")])
```

#### Formatting CPI of product basket

The CSV file also has CPI data over the entire product basket.
We can select the CPI of the product basket from a label.

```{r}
INSEE_CPI_dt_basket <- INSEE_CPI_monthly_values %>%
  .[`Libellé` == "Indice des prix à la consommation - Base 2015 - Ensemble des ménages - France - Ensemble", ] %>%
  # For some stupid reason, INSEE's CSV file has two data series for the same thing.
  .[1]
```

And we basically do the same thing as we did for the CPI per product category.

```{r}
INSEE_CPI_dt_basket <- INSEE_CPI_dt_basket %>%
  melt(id.vars = c("Libellé"),
                   variable.name = "date", value.name = "value") %>%
  .[, .(series_name = `Libellé`,
    year = year(ym(as.character(date))),
    month = month(ym(as.character(date))),
    value = as.numeric(value)
  )]

head(INSEE_CPI_dt[, .SD, .SDcols = !c("series_name")])
```

#### Creating a "cpi" object

Now that we have both the CPI data for individual product categories `INSEE_CPI_dt` and for the entire product basket `INSEE_CPI_dt_basket`, we can initialise a `"cpi"` object that `inflationinequality` functions can read and manipulate.

We would want to create two `"cpi"` objects, one with a maximum COICOP level 2 and another with a maximum COICOP level of 3.
Therefore, we filter `INSEE_CPI_dt` by taking rows with `coicop` codes up to a certain length.

```{r}
INSEE_CPI_dt_level2 <- INSEE_CPI_dt %>% .[nchar(coicop) <= 2 + 1]
INSEE_CPI <- cpi(dt = INSEE_CPI_dt_level2, dt_basket = INSEE_CPI_dt_basket,
                 country = "FR", level = 2)

INSEE_CPI_dt_level3 <- INSEE_CPI_dt %>% .[nchar(coicop) <= 3 + 1]
INSEE_CPI_level3 <- cpi(dt = INSEE_CPI_dt_level3, dt_basket = INSEE_CPI_dt_basket,
                        country = "FR", level = 3)
```

```{r, include=FALSE}
# We need these datasets in another vignette to verify results
saveRDS(INSEE_CPI, "articles/INSEE_CPI_level2.RDS")
saveRDS(INSEE_CPI_level3, "articles/INSEE_CPI_level3.RDS")
```

```{r, include=FALSE}
# We don't include this because it's redundant
annual_values_csv_file <- "famille_IPC-2015/valeurs_annuelles.csv"
INSEE_CPI_annual_values <- as.data.table(read_csv2(annual_values_csv_file))

INSEE_CPI_w_dt <- INSEE_CPI_annual_values %>%
  .[, `:=`(
    idBank = NULL,
    `Dernière mise à jour` = NULL,
    `Période` = NULL
  )] %>%
  .[, coicop := str_extract(
    `Libellé`,
    "(?<=Pondérations de l'indice des prix à la consommation - Base 2015 - Ensemble des ménages - France - Nomenclature Coicop : )\\d{2}(\\.\\d+){0,3}"
    )] %>%
  .[, coicop := str_remove_all(coicop, "\\.")] %>%
  .[!is.na(coicop), ] %>%
  melt(id.vars = c("Libellé", "coicop"), 
                   variable.name = "date", value.name = "weight")  %>%
  .[, `:=`(
    weight = as.numeric(weight),
    year = as.numeric(as.character(date))
  )] %>%
  .[, `:=`(
    date = NULL,
    `Libellé` = NULL
  )]

INSEE_CPI_w_dt_level2 <- INSEE_CPI_w_dt %>% .[nchar(coicop) <= 3]
INSEE_CPI_index_weights <- 
  index_weights(dt = INSEE_CPI_w_dt_level2, country = "FR", level = 2, base_total = 10000)
saveRDS(INSEE_CPI_index_weights, "articles/INSEE_CPI_index_weights_level2.RDS")


INSEE_CPI_w_dt_level3 <- INSEE_CPI_w_dt %>% .[nchar(coicop) <= 4]
INSEE_CPI_index_weights_level3 <- 
  index_weights(dt = INSEE_CPI_w_dt_level3, country = "FR", level = 3, base_total = 10000)
saveRDS(INSEE_CPI_index_weights_level3, "articles/INSEE_CPI_index_weights_level3.RDS")
```

#### Loading HBS data from a XLSX spreadsheet

We can also similarily format HBS data from a XLSX spreadsheet.

```{r}
# Select the file path of the spreadsheet
# Here, we include our example with the package 
XLSX_filepath <- "TF106_3digit.xlsx"

# Read the spreadsheet
TF106_3digit <- read_xlsx(XLSX_filepath)

# (Partially) convert the spreadsheet to our "hbs" format
INSEE_HBS_2017_all <- 
  # Convert to data.table
  as.data.table(TF106_3digit) %>%
  # Convert from wide format to long format
  melt(
    id.vars = c("IDENT", "Regroupement"), variable.name = "quantile",
    value.name = "consumption"
  ) %>%
  # Rearrange columns and remove "Ensemble" (Total) values
  .[, .(
    series_name = Regroupement,
    coicop = IDENT,
    year = 2017,
    category = as.character(quantile),
    consumption
  )] %>%
  # Remove CP00
  .[coicop != "00"]

# Select level 3 COICOPs
INSEE_HBS_2017_dt <- 
  INSEE_HBS_2017_all %>%
  .[nchar(coicop) <= 4 & category != "Ensemble", ]

# Select the total consumption from CP00
# Don't forget to rename consumption to total_consumption!
INSEE_HBS_2017_dt_total <- 
  INSEE_HBS_2017_all %>%
  .[nchar(coicop) <= 4 & category == "Ensemble", 
    .(series_name, coicop, year,
      total_consumption = consumption)]

# Construct the "hbs" object
# Don't forget to specify the order of the categories!
INSEE_HBS_2017 <-
  hbs(dt = INSEE_HBS_2017_dt,
      dt_total = INSEE_HBS_2017_dt_total,
      country = "FR",
      category = "income",
      categories = c("Décile1", "Décile2", "Décile3", "Décile4", "Décile5",
                     "Décile6", "Décile7", "Décile8", "Décile9", "Décile10"),
      level = 3)
```

```{r, include = FALSE}
# Again, for the verification article
saveRDS(INSEE_HBS_2017, "articles/INSEE_HBS_2017_level3.RDS")

INSEE_inflation_income_dt <- INSEE_CPI_annual_values %>%
  .[, `:=`(
    idBank = NULL,
    `Dernière mise à jour` = NULL,
    `Période` = NULL
  )] %>%
  .[`Libellé` %like% "Indice annuel des prix à la consommation - Base 2015 - Niveau de vie du ménage", ] %>%
  melt(id.vars = c("Libellé"), 
                   variable.name = "date", value.name = "cpi")  %>%
  .[!is.na(cpi), .(
    series_name = `Libellé`,
    category = c("Décile2", "Décile3", "Décile4", "Décile5", "Décile6", "Décile7", "Décile8", "Décile9", "Décile10", "Décile1"),
    cpi = as.numeric(cpi),
    year = as.numeric(as.character(date))
  )] %>%
  .[, headline_inflation := ((cpi / shift(cpi, 1)) - 1) * 100, by = .(category)] %>%
  .[!is.na(headline_inflation), .(series_name, category, year, headline_inflation)]

saveRDS(
  INSEE_inflation_income_dt, 
  file = "articles/INSEE_inflation_income_dt.RDS"
)
```

## Loading custom data in a `inflationinequality` function

Since we're using custom datasets, we don't need to specify the country or category when calculating inflation.

```{r}
inflation <- calculate_inflation(
  custom_cpi = INSEE_CPI_level3,
  custom_index_weights = INSEE_CPI_index_weights,
  custom_hbs = INSEE_HBS_2017,
  start_year = 2019
)
plot_time_series(inflation)
```
