---
title: "Using national HBS data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using national HBS data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  warning = FALSE,
  message = FALSE
)
```

```{r setup, include = FALSE, eval = TRUE}
library(inflationinequality)
options(rdbnomics.use_readLines = TRUE)
```

This vignette explains when and how to use national HBS data with `inflationinequality`.

The default workflow should be preferred whenever it is available:

- HICP price indices are downloaded from Eurostat by `load_cpi()`.
- HICP item weights are downloaded from Eurostat by `load_index_weights()`.
- HBS expenditure shares at `level = 1` or `level = 2` are downloaded from Eurostat by `load_hbs()`.

Custom data are mainly useful when a national statistical institute provides HBS expenditure data at a finer COICOP level than Eurostat. In practice, this means using `custom_hbs` for `level = 3` calculations when national HBS data are available. For `level = 2`, the recommended workflow is still the harmonised Eurostat HBS data loaded by `load_hbs()`.

## Recommended workflows

### Level 2: Eurostat HICP and Eurostat HBS

For the harmonised `level = 2` workflow, no custom data are needed. The package downloads the HICP data and the HBS data from Eurostat.

```{r}
inflation <- calculate_inflation("FR", "income", start_year = 2019, level = 2)
plot_time_series(inflation)
```

This is the preferred workflow for cross-country comparisons because the HBS categories are harmonised by Eurostat.

### France level 3: bundled national HBS

France is a special case already implemented in the package. For `country = "FR"`, `level = 3`, and bundled national HBS categories, `inflationinequality` automatically uses INSEE Budget de famille 2017 HBS data when `custom_hbs` is not supplied. HICP prices and HICP weights are still downloaded from Eurostat.

```{r}
inflation_fr_level3 <- calculate_inflation(
  "FR", "income",
  start_year = 2019,
  level = 3
)
plot_time_series(inflation_fr_level3)
```

The same automatic workflow is available for residence-area groups:

```{r}
inflation_fr_urban_level3 <- calculate_inflation(
  "FR", "urban",
  start_year = 2019,
  level = 3
)
plot_time_series(inflation_fr_urban_level3)
```

The automatic workflow is also available for age groups:

```{r}
inflation_fr_age_level3 <- calculate_inflation(
  "FR", "age",
  start_year = 2019,
  level = 3
)
plot_time_series(inflation_fr_age_level3)
```

The INSEE source is available by standard-of-living decile. If a quintile presentation is needed at `level = 3`, use:

```{r}
inflation_fr_quintile <- calculate_inflation(
  "FR", "income",
  start_year = 2019,
  level = 3,
  france_insee_income_groups = "quintile"
)
plot_time_series(inflation_fr_quintile)
```

The option `france_insee_income_groups = "quintile"` averages adjacent INSEE decile mean expenditures (`D1-D2`, `D3-D4`, ..., `D9-D10`). These are expenditure means for equal-population decile groups, not income-threshold bounds.

## When to use custom data

Use custom data when:

- Eurostat HBS is too aggregated for the calculation you want, typically because you need `level = 3`.
- You have national HBS data with the columns needed to build an `"hbs"` object.
- You still want the package to use Eurostat HICP prices and Eurostat HICP item weights.

In that case, provide `custom_hbs` and keep `country`, `category`, and `level` in the calculation call:

```{r}
inflation <- calculate_inflation(
  "FR", "income",
  level = 3,
  start_year = 2019,
  custom_hbs = national_hbs_level3
)
```

You do not need to manually download CPI data from INSEE for the standard package workflow. The package now uses Eurostat HICP downloads for prices and HICP weights. INSEE CPI downloads are only useful for separate validation exercises or for a fully custom CPI workflow.

## Formatting national HBS data

A custom HBS object is built with `hbs()`. It has two data components:

- `dt`: group-specific expenditure by COICOP item.
- `dt_total`: total expenditure by COICOP item.

The `dt` table must contain:

- `series_name`: source label or identifier.
- `coicop`: COICOP code, without the `CP` prefix.
- `year`: HBS survey year.
- `category`: household group, such as income decile or income quintile.
- `consumption`: expenditure amount or share.

The `dt_total` table must contain:

- `coicop`
- `year`
- `total_consumption`

The examples below start from national HBS spreadsheets with expenditure by COICOP and household group. The old BdF scripts such as `2_Calcul_Poids_age.R` and `2_Calcul_Poids_zone_residence.R` merge CPI weights and HBS data month by month. That is no longer necessary for the standard package workflow: `custom_hbs` should contain national HBS expenditure data only, and the package then combines it with Eurostat HICP prices and HICP item weights internally.

```{r}
library(readxl)
library(data.table)
```

The helper below converts a wide national HBS table into an `"hbs"` object. It keeps rows up to COICOP level 3 (`01`, `011`, `0111`, etc.), removes the aggregate `00`, and uses the `Ensemble` column as the all-household total.

```{r}
make_national_hbs_level3 <- function(path, category, group_cols, group_labels,
                                     survey_year = 2017, country = "FR") {
  raw <- as.data.table(read_xlsx(path))

  national_hbs_all <- melt(
    raw,
    id.vars = c("IDENT", "Regroupement", "Ensemble"),
    measure.vars = group_cols,
    variable.name = "source_group",
    value.name = "consumption"
  )

  national_hbs_all[, source_group := as.character(source_group)]
  national_hbs_all[, category_label := group_labels[source_group]]

  national_hbs_dt <- national_hbs_all[
    IDENT != "00" & nchar(IDENT) <= 4,
    .(
      series_name = as.character(Regroupement),
      coicop = as.character(IDENT),
      year = survey_year,
      category = as.character(category_label),
      consumption = as.numeric(consumption)
    )
  ]

  national_hbs_dt_total <- raw[
    IDENT != "00" & nchar(IDENT) <= 4,
    .(
      series_name = as.character(Regroupement),
      coicop = as.character(IDENT),
      year = survey_year,
      total_consumption = as.numeric(Ensemble)
    )
  ]

  hbs(
    dt = national_hbs_dt,
    dt_total = national_hbs_dt_total,
    country = country,
    category = category,
    categories = unname(group_labels),
    level = 3
  )
}
```

### Income groups

`TF106_3digit.xlsx` contains the French national HBS expenditures by standard-of-living decile. The package already bundles this France income `level = 3` source, so this example is mainly useful if you want to reproduce the object explicitly or adapt the workflow to another national source.

```{r}
income_groups <- setNames(
  paste("Decile", 1:10),
  grep("^D", names(read_xlsx("TF106_3digit.xlsx", n_max = 0)), value = TRUE)
)

national_hbs_income_level3 <- make_national_hbs_level3(
  "TF106_3digit.xlsx",
  category = "income",
  group_cols = names(income_groups),
  group_labels = income_groups
)
```

### Residence area groups

`TF104_3digit.xlsx` provides French national HBS expenditures by residence area. It can be used as `custom_hbs` for `category = "urban"` at `level = 3`.

```{r}
urban_groups <- c(
  Rural = "Rural areas",
  Petites_villes = "Small towns",
  Villes_moyennes = "Medium-sized towns",
  Grandes_villes = "Large cities",
  Paris = "Paris"
)

national_hbs_urban_level3 <- make_national_hbs_level3(
  "TF104_3digit.xlsx",
  category = "urban",
  group_cols = names(urban_groups),
  group_labels = urban_groups
)
```

### Age groups

`TF102_3digit.xlsx` provides French national HBS expenditures by age group. The package already bundles this France age `level = 3` source, so this example is mainly useful if you want to reproduce the object explicitly or adapt the workflow to another national source. The category order matters because it controls the order used in plots and gap calculations.

The bundled age source groups `De65a74ans` and `De75ansetplus` into `"65 years or over"` by averaging the two national expenditure profiles.

```{r}
age_groups <- c(
  Moinsde25ans = "Under 25 years",
  De25a34ans = "25-34 years",
  De35a44ans = "35-44 years",
  De45a54ans = "45-54 years",
  De55a64ans = "55-64 years",
  De65a74ans = "65 years or over",
  De75ansetplus = "65 years or over"
)

national_hbs_age_level3 <- make_national_hbs_level3(
  "TF102_3digit.xlsx",
  category = "age",
  group_cols = names(age_groups),
  group_labels = age_groups
)
```

## Using custom HBS in calculations

Once the HBS object is ready, pass it through `custom_hbs`. The package will still download HICP prices and HICP item weights from Eurostat.

```{r}
inflation <- calculate_inflation(
  "FR", "income",
  level = 3,
  start_year = 2019,
  custom_hbs = national_hbs_income_level3
)
plot_time_series(inflation)
```

For residence area or age, change the `category` argument and pass the matching HBS object:

```{r}
inflation_urban <- calculate_inflation(
  "FR", "urban",
  level = 3,
  start_year = 2019,
  custom_hbs = national_hbs_urban_level3
)

inflation_age <- calculate_inflation(
  "FR", "age",
  level = 3,
  start_year = 2019,
  custom_hbs = national_hbs_age_level3
)
```

The same object can be used in lower-level functions:

```{r}
weights <- calculate_weights(
  "FR", "income",
  level = 3,
  start_year = 2019,
  custom_hbs = national_hbs_income_level3
)

indices <- calculate_price_indices(
  "FR", "income",
  level = 3,
  start_year = 2019,
  custom_hbs = national_hbs_income_level3
)
```

If you also provide a fully custom price-index source, use `custom_cpi` and, when needed, `custom_index_weights`. That is a more advanced workflow and should only be used when the Eurostat HICP series are not appropriate for the target calculation.
