---
title: "Verifying calculated inflation"
output: rmarkdown::html_document
vignette: >
  %\VignetteIndexEntry{Verifying calculated inflation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "svg",
  warning = FALSE,
  message = FALSE
)
```

```{r setup, include = FALSE}
library(inflationinequality)

# Vignette specific
library(rdbnomics)
library(data.table)
library(dplyr)
library(lubridate)
library(ggplot2)
library(readr)
library(kableExtra)
```

This vignette calculates the gap between the package's calculated inflation rate and the headline inflation rate published by Eurostat and other national statistical institutes.

***

## Comparing inflation rates in France, Germany and Italy from Eurostat

Our headline inflation will be Eurostat's `prc_hicp_manr` which tracks the monthly annual rate of change of the HICP, and more specifically the "CP00" time series which tracks the entire product basket of a country.

```{r include=FALSE}
load_headline_inflation <- function(country) {
  rdb("Eurostat", "prc_hicp_manr",
                 mask = paste0("M.RCH_A.CP00.", country)
  ) %>%
    .[, .(year = year(period),
          month = month(period),
          coicop,
          headline_inflation = value
    )]

}
```

The calculated inflation rate will be the mean of the inflation rates across all categories, used by `calculate_total_inflation()`.
We will calculate the offset between the calculated inflation and the headline inflation on a month-by-month basis.
The offset is the calculated inflation minus the headline inflation: so if the offset is strictly positive, the calculated inflation is higher than the headline inflation, and vice versa.

```{r include=FALSE}
calculate_inflation_offset <- function(country, category) {
  inflation <- calculate_inflation(country, category)
  dt_total_inflation <- calculate_total_inflation(inflation)
  dt_headline_inflation <- load_headline_inflation(country)

  dt_compared_inflation <- 
    dt_headline_inflation[dt_total_inflation,
                      on = .(year, month)] %>%
    .[, offset := total_inflation - headline_inflation]
}
```

We'll also plot the calculated and headline inflation rates, along side their offset.

```{r, include = FALSE}
plot_inflation_comparison <- function(dt_compared_inflation) {
  scale_factor <- max(dt_compared_inflation$total_inflation, dt_compared_inflation$headline_inflation, na.rm = TRUE) / max(dt_compared_inflation$offset, na.rm = TRUE)
  
  ggplot(dt_compared_inflation) +
    # Time series lines
    geom_line(aes(x = as.Date(paste(year, month, "01", sep = "-")), y = total_inflation, color = "Calculated Inflation")) +
    geom_line(aes(x = as.Date(paste(year, month, "01", sep = "-")), y = headline_inflation, color = "Headline Inflation")) +
    # Histogram bars (scaled)
    geom_col(aes(x = as.Date(paste(year, month, "01", sep = "-")), y = offset * scale_factor), fill = "gray", alpha = 0.6) +
    # Scales and labels
    scale_y_continuous(name = "Inflation Rate",
                       sec.axis = sec_axis(~./scale_factor, name = "Offset")) +
    scale_color_manual(values = c("Calculated Inflation" = "blue", "Headline Inflation" = "red")) +
    labs(title = "Comparison of Calculated and Headline Inflation with Offset",
                  x = "Date",
                  color = "Inflation Type",
                  fill = "Offset") +
    theme_minimal() +
    theme(legend.position = "bottom")
}
```

Let's compare the package's calculated French inflation rate by income with the headline inflation.  

```{r}
dt_compared_inflation_fr_income <- calculate_inflation_offset("FR", "income")
plot_inflation_comparison(dt_compared_inflation_fr_income)
```

By calculating summary statistics of the offset of this inflation data, we can concisely judge how well this inflation data is calculated. 

```{r, include = FALSE}
summarize_inflation_offet_stats <- function(dt_compared_inflation) {
  dt_compared_inflation[, .(
    mean_offset = mean(offset),
    mean_abs_offset = mean(abs(offset)),
    sd_offset = sd(offset),
    sd_abs_offset = sd(abs(offset)),
    min_offset = min(offset),
    max_offset = max(offset)
  )]
}
```

```{r, include = FALSE}
compute_country_diagnostics <- function(country) {
  categories <- c("income", "age", "urban")

  results <- lapply(categories, function(category) {
    dt_compared_inflation <- calculate_inflation_offset(country, category)
    stats <- summarize_inflation_offet_stats(dt_compared_inflation)
    cbind(country = country, category = category, stats)
  })

  # Convert the list to a data.table
  dt_results <- rbindlist(results)

  return(dt_results)
}

compute_diagnostics <- function(countries) {
  results <- lapply(countries, compute_country_diagnostics)

  # Combine all country results into a single data.table
  dt_all_results <- rbindlist(results)

  return(dt_all_results)
}
```

Let's calculate the summary statistics for France, Germany and Italy.

```{r}
inflaton_diagnostics <- compute_diagnostics(c("FR", "DE", "IT"))
knitr::kable(inflaton_diagnostics) %>% kableExtra::kable_styling()
```

## Comparing inflation rates from INSEE

Since INSEE provides more granular inflation and price data, we can use their datasets to more accurately determine the precision of the package's calculated inflation rates.

```{r, include = FALSE}
INSEE_CPI_level3 <- readRDS("INSEE_CPI_level3.RDS")
INSEE_CPI_index_weights_level3 <- readRDS("INSEE_CPI_index_weights_level3.RDS")
INSEE_HBS_2017_level3 <- readRDS("INSEE_HBS_2017_level3.RDS")

insee_inflation_filepath <- "serie_001768594/valeurs_mensuelles.csv"
dt_headline_inflation_insee <- data.table(read_csv2(insee_inflation_filepath)) %>%
  .[, .(year = year(ym(`Libellé`)),
        month = month(ym(`Libellé`)),
        headline_inflation = as.numeric(`Indice des prix à la consommation - Base 2015 - Glissement annuel - Ensemble des ménages - France métropolitaine - Ensemble`))] %>%
  .[complete.cases(.), ]
```

Let's compute the French inflation rates using INSEE's datasets.

```{r}
inflation_insee_level3 <- calculate_inflation("FR", "income", level = 3, custom_cpi = INSEE_CPI_level3, custom_index_weights = INSEE_CPI_index_weights_level3, custom_hbs = INSEE_HBS_2017_level3)
dt_monthly_inflation_insee <- calculate_total_inflation(inflation_insee_level3)
```

Now let's plot the calculated inflation rates and headline rates together with their offset.

```{r, include = FALSE}
dt_compared_inflation <- 
  dt_headline_inflation_insee[dt_monthly_inflation_insee, on = .(year, month)] %>%
  .[, offset := total_inflation - headline_inflation]
```

```{r}
plot_inflation_comparison(dt_compared_inflation)
```

### Comparing annual INSEE inflation by income decile

To provide a more precise comparison, we can examine the annual inflation rate reported by INSEE across different income deciles.

```{r, include = FALSE}
INSEE_inflation_income_dt <- readRDS("INSEE_inflation_income_dt.RDS")
```

We simply calculate the annual inflation rate as the mean of all monthly inflation rates reported in a year.

```{r}
dt_annual_inflation_decile <- inflation_insee_level3$dt[, .(calculated_inflation = mean(inflation)), by = .(year, category)]
```

```{r, include = FALSE}
dt_compared_inflation <-
  INSEE_inflation_income_dt[
    dt_annual_inflation_decile,
    on = .(year, category),
    nomatch = NULL] %>%
  .[, offset := calculated_inflation - headline_inflation]

dt_insee_inflation_stats <- 
  dt_compared_inflation[, .(
    mean_offset = mean(offset),
    mean_abs_offset = mean(abs(offset)),
    sd_offset = sd(offset),
    sd_abs_offset = sd(abs(offset)),
    min_offset = min(offset),
    max_offset = max(offset)
  ), by = .(category)]
```

We obtain the following summary statistics.

```{r}
knitr::kable(dt_insee_inflation_stats) %>% kableExtra::kable_styling()
```

### Comparing level 2 to level 3 aggregation

The more granular the price data, the more accurate the calculated inflation rates are, and that is what we're going to compare below.

```{r, include = FALSE}
INSEE_CPI_level2 <- readRDS("INSEE_CPI_level2.RDS")
INSEE_CPI_index_weights_level2 <- readRDS("INSEE_CPI_index_weights_level2.RDS")
```

We calculate the inflation rates using data from INSEE but at level 2 COICOP.

```{r}
inflation_insee_level2 <- calculate_inflation(country = "FR", category = "income", level = 2, custom_cpi = INSEE_CPI_level2, custom_index_weights = INSEE_CPI_index_weights_level2)
```

```{r, include = FALSE}
dt_inflation_level3 <- 
  calculate_total_inflation(inflation_insee_level3) %>% 
  .[, .(year, month, inflation3 = total_inflation)]
dt_inflation_level2 <- 
  calculate_total_inflation(inflation_insee_level2) %>% 
  .[, .(year, month, inflation2 = total_inflation)]
dt_compared_inflation <- 
  dt_inflation_level3[
    dt_inflation_level2,
    on = .(year, month)
    ] %>%
  .[, offset := inflation3 - inflation2]
```

We obtain the following results.

```{r, echo = FALSE}
# First, calculate the scaling factor
scale_factor <- max(dt_compared_inflation$inflation3, dt_compared_inflation$inflation2, na.rm = TRUE) / max(dt_compared_inflation$offset, na.rm = TRUE)

ggplot(dt_compared_inflation) +
  # Time series lines
  geom_line(aes(x = as.Date(paste(year, month, "01", sep = "-")), y = inflation3, color = "Level 3 Inflation")) +
  geom_line(aes(x = as.Date(paste(year, month, "01", sep = "-")), y = inflation2, color = "Level 2 Inflation")) +
  # Histogram bars (scaled)
  geom_col(aes(x = as.Date(paste(year, month, "01", sep = "-")), y = offset * scale_factor), fill = "gray", alpha = 0.6) +
  # Scales and labels
  scale_y_continuous(name = "Inflation Rate",
                     sec.axis = sec_axis(~./scale_factor, name = "Offset")) +
  scale_color_manual(values = c("Level 3 Inflation" = "blue", "Level 2 Inflation" = "red")) +
  labs(title = "Comparison of Aggregated Inflation with Offset",
       x = "Date",
       color = "Inflation Type",
       fill = "Offset") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Eurozone aggregation

Eurostat also aggregates inflation data across the Eurozone (EA-19).

```{r, include = FALSE}
calculate_weighted_country_inflation <- function(country, category) {
  dt_cow <- rdbnomics::rdb("Eurostat", "prc_hicp_cow",
                 mask = paste0("A.COWEA19.", country)) %>%
    .[, .(year = as.numeric(original_period),
          country_weight = as.numeric(original_value))]
  inflation <- calculate_inflation(country, category)
  dt_total_inflation <- calculate_total_inflation(inflation)
  dt_total_inflation[dt_cow, on = .(year), nomatch = NULL] %>%
    .[, .(country, year, month, category,
          total_inflation,
          country_weight)]
}

aggregate_european_inflation <- function() {
  eurozone_countries <- c("AT", "BE", "CY", "DE", "EE", "EL", "ES", "FI", "IE", "IT", "LT", "LU", "LV", "MT", "NL", "PT", "SI", "SK")
  categories <- c("income", "age", "urban")
  result_list <- list()
  dt_all_result <- data.table::data.table()
  for (category in categories) {
    category_data <- data.table::data.table()
    for (country in eurozone_countries) {
      country_data <- calculate_weighted_country_inflation(country, category)
      category_data <- data.table::rbindlist(list(category_data, country_data))
    }
    euro_inflation <-
      category_data[, .(euro_inflation = weighted.mean(total_inflation,
                                                       country_weight)),
                    by = .(year, month, category)]

    result_list[[category]] <- euro_inflation
    dt_all_result <- data.table::rbindlist(list(dt_all_result, category_data))
  }

  final_result <- data.table::rbindlist(result_list)
  return(list(dt_eurozone = final_result, dt_country = dt_all_result))
}

eurozone_inflation <- aggregate_european_inflation()
dt_headline_eurozone_inflation <-
  rdbnomics::rdb("Eurostat", "prc_hicp_manr",
                 mask = "M.RCH_A.CP00.EA19") %>%
  .[, .(year = lubridate::year(lubridate::ym(original_period)),
        month = lubridate::month(lubridate::ym(original_period)),
        headline_inflation = as.numeric(original_value)
        )]
dt_compared <- dt_headline_eurozone_inflation[eurozone_inflation$dt_eurozone,
                                          on = .(year, month)]

dt_compared_income <-
  summarize_inflation_offet_stats(dt_compared[category == "income", ]) %>%
  .[, category := "income"]
dt_compared_age <-
  summarize_inflation_offet_stats(dt_compared[category == "age", ]) %>%
  .[, category := "age"]
dt_compared_urban <-
  summarize_inflation_offet_stats(dt_compared[category == "urban", ]) %>%
  .[, category := "urban"]
dt_eurozone_offset_stats <- data.table::rbindlist(list(dt_compared_income,
                                                       dt_compared_age,
                                                       dt_compared_urban))
```

Here we compare the package's calculated Eurozone inflation rates with the headline Eurozone inflation rates.

```{r}
knitr::kable(dt_eurozone_offset_stats) %>% kableExtra::kable_styling()
```
