---
title: "Verifying calculated inflation"
output: rmarkdown::html_document
vignette: >
  %\VignetteIndexEntry{Verifying calculated inflation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE
)
```

```{r setup, include = FALSE}
library(inflationinequality)
```

This vignette calculates the gap between the package's calculated inflation rate and the headline inflation rate published by Eurostat and other national statistical institutes.

***

Let's load some libraries.

```{r}
suppressWarnings({
  library(rdbnomics)
  library(data.table)
  library(dplyr)
  library(lubridate)
  library(ggplot2)
})
```


## Comparing inflation rates in France, Germany and Italy

First, let's write a function that downloads the headline inflation rate of a country from Eurostat.

```{r}
load_headline_inflation <- function(country) {
  rdb("Eurostat", "prc_hicp_manr",
                 mask = paste0("M.RCH_A.CP00.", country)
  ) %>%
    .[, .(year = year(period),
          month = month(period),
          coicop,
          headline_inflation = value
    )]

}
```

Now let's write a function that calculates the gap between the headline inflation and the calculated inflation.

```{r}
calculate_inflation_offset <- function(country, category) {
  inflation <- calculate_inflation(country, category)
  dt_total_inflation <- calculate_total_inflation(inflation)
  dt_headline_inflation <- load_headline_inflation(country)

  dt_compared_inflation <- 
    dt_headline_inflation[dt_total_inflation,
                      on = .(year, month)] %>%
    .[, offset := total_inflation - headline_inflation]
}
```

We should also write a plotting function.

```{r}
plot_inflation_comparison <- function(dt_compared_inflation) {
  ggplot(dt_compared_inflation) +
    # Time series lines
    geom_line(aes(x = as.Date(paste(year, month, "01", sep = "-")), y = total_inflation, color = "Calculated Inflation")) +
    geom_line(aes(x = as.Date(paste(year, month, "01", sep = "-")), y = headline_inflation, color = "Headline Inflation")) +
    # Histogram bars
    geom_col(aes(x = as.Date(paste(year, month, "01", sep = "-")), y = offset, fill = "Offset")) +
    # Scales and labels
    scale_y_continuous(name = "Inflation Rate",
                       sec.axis = sec_axis(~., name = "Offset")) +
    scale_color_manual(values = c("Calculated Inflation" = "blue", "Headline Inflation" = "red")) +
    scale_fill_manual(values = c("Offset" = "grey")) +
    labs(title = "Comparison of Calculated and headline Inflation with Offset",
                  x = "Date",
                  color = "Inflation Type",
                  fill = "Offset") +
    theme_minimal() +
    theme(legend.position = "bottom")
}
```

Let's test for French inflation rate by income.

```{r}
dt_compared_inflation_fr_income <- calculate_inflation_offset("FR", "income")
plot_inflation_comparison(dt_compared_inflation_fr_income)
```


We should also get some summary statistics of these offsets.

```{r}
summarize_inflation_offet_stats <- function(dt_compared_inflation) {
  dt_compared_inflation[, .(
    mean_offset = mean(offset),
    mean_abs_offset = mean(abs(offset)),
    sd_offset = sd(offset),
    sd_abs_offset = sd(abs(offset)),
    min_offset = min(offset),
    max_offset = max(offset)
  )]
}
```

And we should compute these statistics for each country and each category.

```{r}
compute_country_diagnostics <- function(country) {
  categories <- c("income", "age", "urban")

  results <- lapply(categories, function(category) {
    dt_compared_inflation <- calculate_inflation_offset(country, category)
    stats <- summarize_inflation_offet_stats(dt_compared_inflation)
    cbind(country = country, category = category, stats)
  })

  # Convert the list to a data.table
  dt_results <- rbindlist(results)

  return(dt_results)
}

compute_diagnostics <- function(countries) {
  results <- lapply(countries, compute_country_diagnostics)

  # Combine all country results into a single data.table
  dt_all_results <- rbindlist(results)

  return(dt_all_results)
}
```

Let's calculate the summary statistics for France, Germany and Italy.

```{r}
inflaton_diagnostics <- compute_diagnostics(c("FR", "DE", "IT"))
```

## Comparing inflation rates to those from INSEE

Let's download CPI data from INSEE.

```{r}
INSEE_CPI <- readRDS("INSEE_CPI_level2.RDS")
INSEE_CPI_index_weights <- readRDS("INSEE_CPI_index_weights_level2.RDS")
```

Now we can compute the inflation rate.

```{r}
inflation_insee <-
    calculate_inflation(
      "FR", "income",
      custom_cpi = INSEE_CPI,
      custom_index_weights = INSEE_CPI_index_weights)
```

Let's write a function that compares headline inflation to calculated inflation from INSEE.

```{r}
dt_monthly_inflation_insee <- calculate_total_inflation(inflation_insee)

insee_inflation_filepath <- "serie_001768594/valeurs_mensuelles.csv"
dt_headline_inflation_insee <- data.table(read_csv2(insee_inflation_filepath)) %>%
  .[, .(year = year(ym(`Libellé`)),
        month = month(ym(`Libellé`)),
        headline_inflation = as.numeric(`Indice des prix à la consommation - Base 2015 - Glissement annuel - Ensemble des ménages - France métropolitaine - Ensemble`))] %>%
  .[complete.cases(.), ]

dt_compared_inflation <-
  dt_headline_inflation_insee[
    dt_monthly_inflation_insee,
    on = .(year, month)] %>%
  .[, offset := total_inflation - headline_inflation]

plot_inflation_comparison(dt_compared_inflation)
```

We also want to load the CPI from INSEE from certain demographic categories.

```{r}
load_insee_inflation <- function(idbank) {
  as.data.table(get_insee_idbank(idbank)) %>%
    .[, .(year = as.numeric(TIME_PERIOD),
          cpi = OBS_VALUE,
          inflation = NA_real_)]
}
```

For income:

```{r}
calculate_cpi_inflation <- function(dt_decile) {
  data.table::setorder(dt_decile, year)
  for (i in 2:nrow(dt_decile)) {
    dt_decile$inflation[i] <- ((dt_decile$cpi[i] / dt_decile$cpi[i-1]) - 1) * 100
  }
  dt_decile
}

compare_insee_quantile_inflation_offset <- function(inflation_insee, idbank1, idbank2, quantile) {
  dt_decile01 <- calculate_cpi_inflation(load_insee_inflation(idbank1))
  dt_decile02 <- calculate_cpi_inflation(load_insee_inflation(idbank2))
  dt_quantile1 <- dt_decile01[dt_decile02, on = .(year)] %>%
    .[!is.na(inflation), .(headline_inflation = mean(inflation)), by = .(year)]

  dt_annual_inflation <-
    inflation_insee$dt[category == inflation_insee$categories[quantile],
                       .(calculated_inflation = mean(inflation)),
                       by = .(year)]
  dt_compared_inflation <-
    dt_quantile1[
      dt_annual_inflation,
      on = .(year),
      nomatch = NULL]
  dt_compared_inflation[, offset := calculated_inflation - headline_inflation]
  dt_compared_inflation
}

dt1 <-
  compare_insee_quantile_inflation_offset(inflation_insee, "010054292", "010054293", 1)
dt2 <-
  compare_insee_quantile_inflation_offset(inflation_insee, "010054294", "010054295", 2)
dt3 <-
  compare_insee_quantile_inflation_offset(inflation_insee, "010054296", "010054297", 3)
dt4 <-
  compare_insee_quantile_inflation_offset(inflation_insee, "010054298", "010054299", 4)
dt5 <-
  compare_insee_quantile_inflation_offset(inflation_insee, "010054300", "010054301", 4)

# Create a list of your data.tables
dt_list <- list(dt1, dt2, dt3, dt4, dt5)
dt_list <- lapply(dt_list, summarize_inflation_offet_stats)

# Combine all data.tables and add the quintile column
combined_dt <- data.table::rbindlist(
  lapply(seq_along(dt_list), function(i) {
    dt_list[[i]][, quintile := inflation_insee$categories[i]]
  })
)

combined_dt
```


## Eurozone aggregation

We can also compare with the Eurozone aggregate.
Beware this code chunk takes more than 30 minutes to compute.

```{r, eval = FALSE}
calculate_weighted_country_inflation <- function(country, category) {
  dt_cow <- rdbnomics::rdb("Eurostat", "prc_hicp_cow",
                 mask = paste0("A.COWEA19.", country)) %>%
    .[, .(year = as.numeric(original_period),
          country_weight = as.numeric(original_value))]
  inflation <- calculate_inflation(country, category)
  dt_total_inflation <- calculate_total_inflation(inflation)
  dt_total_inflation[dt_cow, on = .(year), nomatch = NULL] %>%
    .[, .(country, year, month, category,
          total_inflation,
          country_weight)]
}

aggregate_european_inflation <- function() {
  eurozone_countries <- c("AT", "BE", "CY", "DE", "EE", "EL", "ES", "FI", "IE", "IT", "LT", "LU", "LV", "MT", "NL", "PT", "SI", "SK")
  categories <- c("income", "age", "urban")
  result_list <- list()
  dt_all_result <- data.table::data.table()
  for (category in categories) {
    category_data <- data.table::data.table()
    for (country in eurozone_countries) {
      country_data <- calculate_weighted_country_inflation(country, category)
      category_data <- data.table::rbindlist(list(category_data, country_data))
    }
    euro_inflation <-
      category_data[, .(euro_inflation = weighted.mean(total_inflation,
                                                       country_weight)),
                    by = .(year, month, category)]

    result_list[[category]] <- euro_inflation
    dt_all_result <- data.table::rbindlist(list(dt_all_result, category_data))
  }

  final_result <- data.table::rbindlist(result_list)
  return(list(dt_eurozone = final_result, dt_country = dt_all_result))
}

eurozone_inflation <- aggregate_european_inflation()
dt_headline_eurozone_inflation <-
  rdbnomics::rdb("Eurostat", "prc_hicp_manr",
                 mask = "M.RCH_A.CP00.EA19") %>%
  .[, .(year = lubridate::year(lubridate::ym(original_period)),
        month = lubridate::month(lubridate::ym(original_period)),
        headline_inflation = as.numeric(original_value)
        )]
dt_compared <- dt_headline_eurozone_inflation[eurozone_inflation$dt_eurozone,
                                          on = .(year, month)]

dt_compared_income <-
  summarize_inflation_offet_stats(dt_compared[category == "income", ]) %>%
  .[, category := "income"]
dt_compared_age <-
  summarize_inflation_offet_stats(dt_compared[category == "age", ]) %>%
  .[, category := "age"]
dt_compared_urban <-
  summarize_inflation_offet_stats(dt_compared[category == "urban", ]) %>%
  .[, category := "urban"]
dt_eurozone_offset_stats <- data.table::rbindlist(list(dt_compared_income,
                                                       dt_compared_age,
                                                       dt_compared_urban))
```

## Comparing aggregation at COICOP level 2 and level 3

Level 3 CPI value data

```{r}
INSEE_CPI_dt3 <- INSEE_CPI_monthly_values %>%
  .[, `:=`(
    idBank = NULL,
    `Dernière mise à jour` = NULL,
    `Période` = NULL
  )] %>%
  .[, coicop := str_extract(
    `Libellé`,
    "(?<=Indice des prix à la consommation - Base 2015 - Ensemble des ménages - France - Nomenclature Coicop : )\\d{2}(\\.\\d+){0,3}"
    )] %>%
  .[, coicop := str_remove_all(coicop, "\\.")] %>%
  melt(id.vars = c("Libellé", "coicop"), 
                   variable.name = "date", value.name = "value") %>%
  .[nchar(coicop) == 4] %>%
  .[, `:=`(
    date = as.character(date),
    value = as.numeric(value),
    series_name = `Libellé`,
    year = year(ym(date)),
    month = month(ym(date))
  )] %>%
  .[, `:=`(
    date = NULL,
    `Libellé` = NULL
  )]

INSEE_CPI3 <- cpi(dt = INSEE_CPI_dt3, dt_basket = INSEE_CPI_dt_basket,
                 country = "FR", level = 3)
```

Level 3 CPI weights

```{r}
INSEE_CPI_w_dt3 <- INSEE_CPI_annual_values %>%
  .[, `:=`(
    idBank = NULL,
    `Dernière mise à jour` = NULL,
    `Période` = NULL
  )] %>%
  .[, coicop := str_extract(
    `Libellé`,
    "(?<=Pondérations de l'indice des prix à la consommation - Base 2015 - Ensemble des ménages - France - Nomenclature Coicop : )\\d{2}(\\.\\d+){0,3}"
    )] %>%
  .[, coicop := str_remove_all(coicop, "\\.")] %>%
  melt(id.vars = c("Libellé", "coicop"), 
                   variable.name = "date", value.name = "weight") %>%
  .[nchar(coicop) == 4] %>%
  .[, `:=`(
    weight = as.numeric(weight),
    year = as.numeric(as.character(date))
  )] %>%
  .[, `:=`(
    date = NULL,
    `Libellé` = NULL
  )]

INSEE_CPI_index_weights3 <- 
  index_weights(dt = INSEE_CPI_w_dt3, country = "FR", level = 3)
```

Level 3 HBS

```{r}
# Select the file path of the spreadsheet
# Here, we include our example with the package 
XLSX_filepath <- 
  system.file("extdata", "TF106_3digit.xlsx", package = "inflationinequality")

# Read the spreadsheet
library(readxl)
TF106_3digit <- read_xlsx(XLSX_filepath)

# (Partially) convert the spreadsheet to our "hbs" format
INSEE_HBS_2017_all <- 
  # Convert to data.table
  as.data.table(TF106_3digit) %>%
  # Convert from wide format to long format
  melt(
    id.vars = c("IDENT", "Regroupement"), variable.name = "quantile",
    value.name = "consumption"
  ) %>%
  # Rearrange columns and remove "Ensemble" (Total) values
  .[, .(
    series_name = Regroupement,
    coicop = IDENT,
    year = 2017,
    category = quantile,
    consumption
  )]

# Select level 3 COICOPs
INSEE_HBS_2017_dt <- 
  INSEE_HBS_2017_all %>%
  .[nchar(coicop) == 4 & category != "Ensemble", ]

# Select the total consumption from CP00
# Don't forget to rename consumption to total_consumption!
INSEE_HBS_2017_dt_total <- 
  INSEE_HBS_2017_all %>%
  .[nchar(coicop) == 4 & category == "Ensemble", 
    .(series_name, coicop, year,
      total_consumption = consumption)]

# Construct the "hbs" object
# Don't forget to specify the order of the categories!
INSEE_HBS_2017 <-
  hbs(dt = INSEE_HBS_2017_dt,
      dt_total = INSEE_HBS_2017_dt_total,
      country = "FR",
      category = "income",
      categories = c("Décile1", "Décile2", "Décile3", "Décile4", "Décile5",
                     "Décile6", "Décile7", "Décile8", "Décile9", "Décile10"),
      level = 3)
```

Calculate inflation at level 3 COICOP

```{r}
inflation_fr3 <- calculate_inflation(country = "FR", category = "income", level = 3,
                                     custom_cpi = INSEE_CPI3,
                                     custom_index_weights = INSEE_CPI_index_weights3,
                                     custom_hbs = INSEE_HBS_2017)
```
```{r}

```
