---
title: "Verifying calculated inflation"
output: rmarkdown::html_document
vignette: >
  %\VignetteIndexEntry{Verifying calculated inflation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE
)
```

```{r setup, include = FALSE}
library(inflationinequality)
```

This vignette calculates the gap between the package's calculated inflation rate and the headline inflation rate published by Eurostat and other national statistical institutes.

***

Let's load some libraries.

```{r}
suppressWarnings({
  library(rdbnomics)
  library(data.table)
  library(dplyr)
  library(lubridate)
  library(ggplot2)
})
```


## Comparing inflation rates in France, Germany and Italy

First, let's write a function that downloads the headline inflation rate of a country from Eurostat.

```{r}
load_headline_inflation <- function(country) {
  rdb("Eurostat", "prc_hicp_manr",
                 mask = paste0("M.RCH_A.CP00.", country)
  ) %>%
    .[, .(year = year(period),
          month = month(period),
          coicop,
          headline_inflation = value
    )]

}
```

Now let's write a function that calculates the gap between the headline inflation and the calculated inflation.

```{r}
calculate_inflation_offset <- function(country, category) {
  inflation <- calculate_inflation(country, category)
  dt_total_inflation <- calculate_total_inflation(inflation)
  dt_headline_inflation <- load_headline_inflation(country)

  dt_compared_inflation <- 
    dt_headline_inflation[dt_total_inflation,
                      on = .(year, month)] %>%
    .[, offset := total_inflation - headline_inflation]
}
```

We should also write a plotting function.

```{r}
plot_inflation_comparison <- function(dt_compared_inflation) {
  scale_factor <- max(dt_compared_inflation$total_inflation, dt_compared_inflation$headline_inflation, na.rm = TRUE) / max(dt_compared_inflation$offset, na.rm = TRUE)
  
  ggplot(dt_compared_inflation) +
    # Time series lines
    geom_line(aes(x = as.Date(paste(year, month, "01", sep = "-")), y = total_inflation, color = "Calculated Inflation")) +
    geom_line(aes(x = as.Date(paste(year, month, "01", sep = "-")), y = headline_inflation, color = "Headline Inflation")) +
    # Histogram bars (scaled)
    geom_col(aes(x = as.Date(paste(year, month, "01", sep = "-")), y = offset * scale_factor), fill = "gray", alpha = 0.6) +
    # Scales and labels
    scale_y_continuous(name = "Inflation Rate",
                       sec.axis = sec_axis(~./scale_factor, name = "Offset")) +
    scale_color_manual(values = c("Calculated Inflation" = "blue", "Headline Inflation" = "red")) +
    labs(title = "Comparison of Calculated and Headline Inflation with Offset",
                  x = "Date",
                  color = "Inflation Type",
                  fill = "Offset") +
    theme_minimal() +
    theme(legend.position = "bottom")
}
```

Let's test for French inflation rate by income.

```{r}
dt_compared_inflation_fr_income <- calculate_inflation_offset("FR", "income")
plot_inflation_comparison(dt_compared_inflation_fr_income)
```


We should also get some summary statistics of these offsets.

```{r}
summarize_inflation_offet_stats <- function(dt_compared_inflation) {
  dt_compared_inflation[, .(
    mean_offset = mean(offset),
    mean_abs_offset = mean(abs(offset)),
    sd_offset = sd(offset),
    sd_abs_offset = sd(abs(offset)),
    min_offset = min(offset),
    max_offset = max(offset)
  )]
}
```

And we should compute these statistics for each country and each category.

```{r}
compute_country_diagnostics <- function(country) {
  categories <- c("income", "age", "urban")

  results <- lapply(categories, function(category) {
    dt_compared_inflation <- calculate_inflation_offset(country, category)
    stats <- summarize_inflation_offet_stats(dt_compared_inflation)
    cbind(country = country, category = category, stats)
  })

  # Convert the list to a data.table
  dt_results <- rbindlist(results)

  return(dt_results)
}

compute_diagnostics <- function(countries) {
  results <- lapply(countries, compute_country_diagnostics)

  # Combine all country results into a single data.table
  dt_all_results <- rbindlist(results)

  return(dt_all_results)
}
```

Let's calculate the summary statistics for France, Germany and Italy.

```{r}
inflaton_diagnostics <- compute_diagnostics(c("FR", "DE", "IT"))
```

## Comparing inflation rates to those from INSEE

Let's download CPI data from INSEE.

```{r}
INSEE_CPI <- readRDS("INSEE_CPI_level2.RDS")
INSEE_CPI_index_weights <- readRDS("INSEE_CPI_index_weights_level2.RDS")
```

Now we can compute the inflation rate.

```{r}
contributions_insee <-
  calculate_contributions(
    "FR", "income",
    custom_cpi = INSEE_CPI,
    custom_index_weights = INSEE_CPI_index_weights)

inflation_insee <-
    calculate_inflation(
      "FR", "income",
      custom_cpi = INSEE_CPI,
      custom_index_weights = INSEE_CPI_index_weights)
```

Let's write a function that compares headline inflation to calculated inflation from INSEE.

```{r}
dt_monthly_inflation_insee <- calculate_total_inflation(inflation_insee)

insee_inflation_filepath <- "serie_001768594/valeurs_mensuelles.csv"
dt_headline_inflation_insee <- data.table(read_csv2(insee_inflation_filepath)) %>%
  .[, .(year = year(ym(`Libellé`)),
        month = month(ym(`Libellé`)),
        headline_inflation = as.numeric(`Indice des prix à la consommation - Base 2015 - Glissement annuel - Ensemble des ménages - France métropolitaine - Ensemble`))] %>%
  .[complete.cases(.), ]

dt_compared_inflation <-
  dt_headline_inflation_insee[
    dt_monthly_inflation_insee,
    on = .(year, month)] %>%
  .[, offset := total_inflation - headline_inflation]

plot_inflation_comparison(dt_compared_inflation)
```

## Eurozone aggregation

We can also compare with the Eurozone aggregate.
Beware this code chunk takes more than 30 minutes to compute.

```{r, eval = FALSE}
calculate_weighted_country_inflation <- function(country, category) {
  dt_cow <- rdbnomics::rdb("Eurostat", "prc_hicp_cow",
                 mask = paste0("A.COWEA19.", country)) %>%
    .[, .(year = as.numeric(original_period),
          country_weight = as.numeric(original_value))]
  inflation <- calculate_inflation(country, category)
  dt_total_inflation <- calculate_total_inflation(inflation)
  dt_total_inflation[dt_cow, on = .(year), nomatch = NULL] %>%
    .[, .(country, year, month, category,
          total_inflation,
          country_weight)]
}

aggregate_european_inflation <- function() {
  eurozone_countries <- c("AT", "BE", "CY", "DE", "EE", "EL", "ES", "FI", "IE", "IT", "LT", "LU", "LV", "MT", "NL", "PT", "SI", "SK")
  categories <- c("income", "age", "urban")
  result_list <- list()
  dt_all_result <- data.table::data.table()
  for (category in categories) {
    category_data <- data.table::data.table()
    for (country in eurozone_countries) {
      country_data <- calculate_weighted_country_inflation(country, category)
      category_data <- data.table::rbindlist(list(category_data, country_data))
    }
    euro_inflation <-
      category_data[, .(euro_inflation = weighted.mean(total_inflation,
                                                       country_weight)),
                    by = .(year, month, category)]

    result_list[[category]] <- euro_inflation
    dt_all_result <- data.table::rbindlist(list(dt_all_result, category_data))
  }

  final_result <- data.table::rbindlist(result_list)
  return(list(dt_eurozone = final_result, dt_country = dt_all_result))
}

eurozone_inflation <- aggregate_european_inflation()
dt_headline_eurozone_inflation <-
  rdbnomics::rdb("Eurostat", "prc_hicp_manr",
                 mask = "M.RCH_A.CP00.EA19") %>%
  .[, .(year = lubridate::year(lubridate::ym(original_period)),
        month = lubridate::month(lubridate::ym(original_period)),
        headline_inflation = as.numeric(original_value)
        )]
dt_compared <- dt_headline_eurozone_inflation[eurozone_inflation$dt_eurozone,
                                          on = .(year, month)]

dt_compared_income <-
  summarize_inflation_offet_stats(dt_compared[category == "income", ]) %>%
  .[, category := "income"]
dt_compared_age <-
  summarize_inflation_offet_stats(dt_compared[category == "age", ]) %>%
  .[, category := "age"]
dt_compared_urban <-
  summarize_inflation_offet_stats(dt_compared[category == "urban", ]) %>%
  .[, category := "urban"]
dt_eurozone_offset_stats <- data.table::rbindlist(list(dt_compared_income,
                                                       dt_compared_age,
                                                       dt_compared_urban))
```

## Comparing aggregation at COICOP level 2 and level 3

Level 3 CPI value data

```{r}
INSEE_CPI_level3 <- readRDS("INSEE_CPI_level3.RDS")
```

Level 3 CPI weights

```{r}
INSEE_CPI_index_weights_level3 <- 
  readRDS("INSEE_CPI_index_weights_level3.RDS")
```

Level 3 HBS

```{r}
INSEE_HBS_2017 <-
  readRDS("INSEE_HBS_2017_level3.RDS")
```

Calculate inflation at level 3 COICOP

```{r}
inflation_level3 <- calculate_inflation(country = "FR", category = "income", level = 3,
                                     custom_cpi = INSEE_CPI_level3,
                                     custom_index_weights = INSEE_CPI_index_weights_level3,
                                     custom_hbs = INSEE_HBS_2017)
```

```{r}
dt_inflation_level3 <- 
  calculate_total_inflation(inflation_level3) %>% 
  .[, .(year, month, inflation3 = total_inflation)]
dt_inflation_level2 <- 
  calculate_total_inflation(inflation_insee) %>% 
  .[, .(year, month, inflation2 = total_inflation)]
dt_compared_inflation <- 
  dt_inflation_level3[
    dt_inflation_level2,
    on = .(year, month)
    ] %>%
  .[, offset := inflation3 - inflation2]

# First, calculate the scaling factor
scale_factor <- max(dt_compared_inflation$inflation3, dt_compared_inflation$inflation2, na.rm = TRUE) / max(dt_compared_inflation$offset, na.rm = TRUE)

ggplot(dt_compared_inflation) +
  # Time series lines
  geom_line(aes(x = as.Date(paste(year, month, "01", sep = "-")), y = inflation3, color = "Level 3 Inflation")) +
  geom_line(aes(x = as.Date(paste(year, month, "01", sep = "-")), y = inflation2, color = "Level 2 Inflation")) +
  # Histogram bars (scaled)
  geom_col(aes(x = as.Date(paste(year, month, "01", sep = "-")), y = offset * scale_factor), fill = "gray", alpha = 0.6) +
  # Scales and labels
  scale_y_continuous(name = "Inflation Rate",
                     sec.axis = sec_axis(~./scale_factor, name = "Offset")) +
  scale_color_manual(values = c("Level 3 Inflation" = "blue", "Level 2 Inflation" = "red")) +
  labs(title = "Comparison of Aggregated Inflation with Offset",
       x = "Date",
       color = "Inflation Type",
       fill = "Offset") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Comparing INSEE inflation by income quintile


We also want to load the CPI from INSEE from certain demographic categories.

Load headline inflation and calculate inflation from level 3 CPI and HBS data from INSEE.

```{r}
INSEE_inflation_income_dt <-
  readRDS("INSEE_inflation_income_dt.RDS")

inflation_fr_income3 <-
  calculate_inflation(
    "FR", "income", level = 3,
    custom_cpi = INSEE_CPI_level3,
    custom_index_weights = INSEE_CPI_index_weights_level3,
    custom_hbs = INSEE_HBS_2017)

```

We compare each income decile's annualized inflation rate and compute summary statistics.

```{r}
dt_annual_inflation_decile <-
  inflation_fr_income3$dt[, .(calculated_inflation = mean(inflation)), by = .(year, category)]

dt_compared_inflation <-
  INSEE_inflation_income_dt[
    dt_annual_inflation_decile,
    on = .(year, category),
    nomatch = NULL] %>%
  .[, offset := calculated_inflation - headline_inflation]

dt_insee_inflation_stats <- 
  dt_compared_inflation[, .(
    mean_offset = mean(offset),
    mean_abs_offset = mean(abs(offset)),
    sd_offset = sd(offset),
    sd_abs_offset = sd(abs(offset)),
    min_offset = min(offset),
    max_offset = max(offset)
  ), by = .(category)]

dt_insee_inflation_stats
```


